Causal Forest Approach for Site-specific Input Management via On-farm Precision Experimentation

### 2022-02-06 Writing/Re_JournalSubmission/ML_VRA_v2.rmd
rewriting the second section

# Background:Using Causal Forest Methods to Predict Changes in Yield

In the following, we use basic microeconomic theory to describe the fundamental information necessary to identify EOIRs in general: the site-specific treatment effect of input application on yield. We then discuss the causal forest method in detail, show how it is particularly suitable to obtain this fundamental information, and contrast it to more traditional prediction-oriented ML methods, such as RF, BRF, and CNN.

## A Simple Profit-maximization Problem

@@

Elementary calculus shows that the necessary condition for maximization is, that $\partial f(I, \mathbf{c})/\partial I - w =0$, which implicitly defines the economically optimal input application rate as a function $I^{*}$ of prices and soil/field characteristics, and may be arranged as \@ref(eq:eqn2):

@@

Identity \@ref(eq:eqn2) states that the necessary elements of the information needed for profit maximization are the input price, the output price, and the partial derivative of the yield function with respect to the input. While knowledge of the yield function $f(I, \mathbf{c})$ in Eq. \@ref(eq:eqn1) is sufficient to solve the maximization problem above (since its derivative can be calculated), identity \@ref(eq:eqn2) shows that it is not a necessary condition; rather, knowing the function's partial derivative is necessary.

In the following, we discuss a method of using Causal Forest methods that can provide statistical advantages to the estimation of economically optimal input application rates because it is built to estimate approximations of the derivative $\partial f/\partial I$ directly, and not the yield function $f$ itself.

Recent studies using ML methods to estimate the effectiveness of various agricultural input management strategies have based their analyses on estimates of a yield function, $f(I, \mathbf{c})$ (e.g., @krause2020random; @barbosa2020modeling; @Gardner21). In the following, we present a Causal Forest methodology that does not rely on estimation of how the level of yield depends on the input application rate $I$, but rather relies on a discrete version of Eq. \@ref(eq:eqn2), examining the discrete changes in yield that result from discrete *changes* in the input application rate $I$.

Consider an increasing ordering of input application rates $I_1, I_2,...,I_K$. Among these K rates, for a generic site $i$ a specific EOIR can be identified if the *treatment effect* function $\Delta y_i(I_{k+1}, I_{k})$ defined in Eq. \@ref(eq:eqn3)is known. This function is defined as the discrete change in yield that results from a discrete change in the input level from $I_k$ to $I_{k+1}$, and it depends on the characteristics at the site analyzed:

\begin{equation}
\Delta y_i(I_{k+1}, I_{k}) = f(I_{k+1}, \mathbf{c_i}) - f(I_{k}; \mathbf{c_i}) \quad ^\forall k = 1, \dots , K - 1. (\#eq:eqn3)
\end{equation}

Once treatment effects $\Delta y_i(I_{k+1}, I_{k})$ have been identified for every $k$, economic analysis can be conducted to identify the EOIR for site $i$. The profit change from increasing the input rate from $I_{1}$ to $I_{k}$ is

\begin{equation}
p (\sum_1^{k-1} \Delta y_i(I_{k+1}, I_{k})) - w \times (I_k - I_0). (\#eq:eqn4)
\end{equation}

The EOIR is the value $I_k$ that maximizes the profit gain "relative" to profits from $I_1$. ($I_1$ may be the EOIR if no other rates generates greater profit.)

@@

```{r ct-ex, out.width="50%", fig.align="center", fig.cap = "An illustrative example of causal tree"}
# knitr::include_graphics("./GitControlled/Writing/Re_JournalSubmission/ctree_ex.pdf")
readPNG("./GitControlled/Writing/Re_JournalSubmission/ctree_ex.png") %>% grid.raster()
```

@@

The first data split criterion is whether the $\textit{ecs}$ value is greater or less than $25$. The observations with $\textit{ecs}$ greater than or equal to $25$ move to the left node, while those with $\textit{ecs}$ less than $25$ move to the right. For each of the subsamples, the treatment effect is identified as the difference in the mean yields of the treated and control groups within the subsample. The treatment effect for the low-$\textit{ecs}$ observations was $619$ kg/ha, and for the high-$\textit{ecs}$ observations was $982$ kg/ha. This means that for the impact of the increase in seed rate the treatment effect is heterogeneous over space, having a larger positive impact on yield in high-$\textit{ecs}$ areas. Trees are further developed to have more groups of subsamples to allow for more flexible representations of the heterogeneous treatment effects. The estimated CF model does not predict yield level at any site $i$ in the field, but rather estimates site-specific EOIRs by estimating treatment effects. Since CF is designed to estimate the causal effect of an input directly, it may be suspected that CF will outperform other ML methods that identify EOIR indirectly through yield prediction.

@@

## Advantages of Causal Forest Analysis in the Context of On-farm Precision Experimentation

Whether the indirect (yield-prediction-first) or direct approach is used can lead to meaningful statistical differences in EOIR estimation. The conventional prediction-oriented ML methods are designed to predict yield, $y_{i}(I) \equiv f(I, c_{i})$, well, but not necessarily treatment effects. Since the loss function used in prediction-oriented ML uses (typically squared) residuals in yield prediction, prediction-oriented ML is naturally optimized for prediction the "level" of yield instead of identifying the causal impact of the input. But CF is designed specifically to estimate the casual impact of the input. Indeed, @Wager2018a used Monte Carlo simulations to show that CF provided more accurate estimates of heterogeneous treatment effects than those provided by RF. However, the underlying data generating process of their simulations was extremely simple, with the treatment effect influenced by only a single variable. CF has never been tested for more complicated data generating processes, such as those represented by yield response functions in which the treatment effect of an input can be affected by multiple variables in a non-linear fashion. Moreover, CF has never been compared to more advanced ML methods such as BRF and CNN that are considered to have prediction capabilities greater than RF's. Our report builds on @Wager2018a to compare CF to these prediction-oriented ML methods (not just RF, but also BRF and CNN) in more realistic agronomic experiment settings.

>>>>

# Predicting Yield vs. Predicting Changes in Yield in Response to Changes in Input Management

In the following, we describe the fundamental information necessary to identify EOIR in general: the site-specific treatment effect of input application on yield. We then discuss the causal forest method in detail, show how it is particularly suitable to obtain this fundamental information, and contrast it to more traditional prediction-oriented ML methods, such as RF, BRF, and CNN.

@@

Elementary calculus shows that the necessary condition for maximization is, that $\partial f(I, \mathbf{c})/\partial I - w =0$, which implicitly defines the economically optimal input application rate as a function of prices and soil/field characteristics, and may be arranged as \@ref(eq:eqn2):

@@

Identity \@ref(eq:eqn2) states that the necessary elements of the information needed for profit maximization are the input price, the output price, and the partial derivative of the yield function with respect to the input. While knowledge of the yield function $f(I, \mathbf{c})$ in Eq. \@ref(eq:eqn1)) is sufficient to solve the maximization problem above (since its derivative can be calculated), identity \@ref(eq:eqn2) shows that it is not a necessary condition; rather, knowing the _change_ in yield in response to a change in input rate is.

Recent papers using ML methods to estimate the effectiveness of various agricultural input management strategies first estimate yield function, $f(I, \mathbf{c})$ (e.g., @krause2020random; @barbosa2020modeling; @Gardner21) and then derive changes in yield associated with changes in input rate to estimate site-specific input recommendations. On the other hand, Causal Forest is designed specifically to estimate a discrete version of Eq. \@ref(eq:eqn2) directly, examining the discrete changes in yield that result from discrete *changes* in the input application rate $I$, bypassing the estimation of $f(I, \mathbf{c})$. It may be suspected that CF will outperform other ML methods that identify EOIR indirectly through yield prediction. We now details how CF works below and why CF may perform better in predicting EOIR than other prediction-oriented ML methods used in the previous studies.

@@

```{r ct-ex, fig.width=4, fig.height=4, fig.align="center", fig.cap = "An illustrative example of causal tree"}
# knitr::include_graphics("./GitControlled/Writing/Re_JournalSubmission/ctree_ex.pdf")
readPNG("./GitControlled/Writing/Re_JournalSubmission/ctree_ex.png") %>% grid.raster()
```

@@

The first data split criterion is whether the $\textit{ecs}$ value is greater or less than $25$. The observations with $\textit{ecs}$ greater than or equal to $25$ move to the left node, while those with $\textit{ecs}$ less than $25$ move to the right. For each of the subsamples, the treatment effect is identified as the difference in the mean yields of the treated and control groups within the subsample. The treatment effect for the low-$\textit{ecs}$ observations was $619$ kg/ha, and for the high-$\textit{ecs}$ observations was $982$ kg/ha. This means that for the impact of the increase in seed rate the treatment effect is heterogeneous over space, having a larger positive impact on yield in high-$\textit{ecs}$ areas. Trees are further developed to have more groups of subsamples to allow for more flexible representations of the heterogeneous treatment effects. The estimated heterogeneous treatments effects can then be used to identify site-specific EOIRs. It is noteworth that at no point CF estimates yield _level_.

@@

## CF vs. Prediction-oriented ML methods

Whether the indirect (yield-prediction-first) and direct approach is used can lead to meaningful statistical differences when it comes to EOIR estimation. The conventional prediction-oriented ML methods are designed to predict yield, $y_i(I,\mathbf{c})$, well, but not necessarily treatment effects. Since the loss function used in prediction-oriented ML uses (typically squared) residuals in yield prediction, prediction-oriented ML is naturally optimized for predicting the "level" of yield instead of identifying the causal impact of the input on yield. CF, on the other hand, is designed specifically to estimate the causal impact of input. Indeed, @Wager2018a used Monte Carlo simulations to show that CF provided more accurate estimates of heterogeneous treatment effects than those provided by RF. However, the underlying data generating process of their simulations was extremely simple, with the treatment effect influenced by only a single variable. CF has not been tested for more complicated data generating processes, such as those represented by yield response functions in which the treatment effect of an input can be affected by multiple variables in a non-linear fashion. Moreover, CF has never been compared to more advanced ML methods such as BRF and CNN that are considered to have prediction capabilities greater than RF's. The study reported here builds on @Wager2018a to compare CF to these prediction-oriented ML methods (not just RF, but also BRF and CNN) in more realistic agronomic experiment settings.
